---
title: "Drawing log-data"
author: "Tomasz Żółtak"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Drawing log-data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6
)
```

# Data and setup

We will use the *arrow* library to read large CSV files into R and *dplyr* to perform some data manipulation.

```{r}
library(logLime)
library(arrow, warn.conflicts = FALSE)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
library(ggdensity)
```

Data we will use are of two types:

1.  Survey results exported from Lime Survey platform to a CSV file (encoded in UTF-8 with BOM).
    -   In fact the file was created by combining two such files from different surveys.
    -   We need to read this file to R by ourselves - below it is be imported into a data frame named `surveyResults`.
        - Function from the *arrow* library is used below to speed up reading the file, but please note that non-default settings are needed to be set to enable reading data containing such long records.
2.  Two survey structure files exported to TXT files (tab-separated).
    -   There are two such files because results come from two different surveys.
    -   We only need to provide names (locations) of these files, as *logLime*'s function may handle reading these files internally.

```{r}
surveyResusultFile <- "https://tomek.zozlak.org/dataExp5BOM.csv"
surveyStructurFiles <- c("https://tomek.zozlak.org/limesurvey_survey_745195.txt",
                         "https://tomek.zozlak.org/limesurvey_survey_817417.txt")

timeoutDefault <- getOption("timeout")
options(timeout = Inf)
surveyResults <- read_csv_arrow(url(surveyResusultFile),
                                read_options =
                                  CsvReadOptions$create(block_size = 10^8))
options(timeout = timeoutDefault)
```

Having the survey results (along wit log-data) read into R code below perform preprocessing, filter out respondent-screens with some common problems identified, computes and adds columns with *relative* positions and distances and finally merge some respondent's characteristics to the data regarding *actions* (events).

```{r}
# creating log-data object
logData <- separate_logdata_types(surveyResults %>%
                                    select(respid, starts_with("log")),
                                  surveyStructure = surveyStructurFiles,
                                  questionNamesPrefix = "log")
# filtering out problematic records (respondent-screens)
logData <- lapply(logData, semi_join,
                  y = logData$systemInfo %>%
                    filter(problemsAnyBroken != 1,
                           problemsLeftBrowser != 1,
                           problemsResized != 1,
                           problemsTimeStamps != 1,
                           inputsWidth > 0),
                  by = c("respid", "screen"))
# computing relative positions and distances
logData$actions <- compute_relative_positions(logData$actions,
                                              logData$systemInfo)
# merging some respondent's characteristics
logData$actions <- logData$actions %>%
  left_join(surveyResults %>%
              select(respid, variant, gender),
            by = "respid")
```

Now we may proceed to drawing.

# Drawing heatmaps

One of the commonly used form of visualization of cursor moves data is preparing so-called heat-maps, marking areas of the survey screen that the cursor was pointing at for a long time and areas that were hardly visited (pointed at). Interesting feature of this form of visualization is that it can be used both to illustrate actions of a single respondent and to illustrate aggregated tendencies among group of respondents.

## Data preparation

One may draw heat-maps using function `geom_density2d_filled()` from the *ggplot2* package. However, data prepared by the `separate log data types()` above can not (or, more specifically, should not) be used to create such plots. That's because the *actions* element of a log-data objects contain *mousemove* events of different duration, and while estimating time that pointer was at a given position these duration should be used a weights. But there is no way to provide different weights while computing 2-dimensional densities in *ggplot2*. Because of this data need to be transformed to a different form in which cursor positions will be reported across evenly spaced time points. You may get data of this form by using function `compute_cursor_positions()`:

```{r}
(cursorPositions <- compute_cursor_positions(logData$actions,
                                             timeSpan = 250) %>%
   # also, let's join this data with some respondent's characteristics
   left_join(surveyResults %>%
               select(respid, variant, gender),
             by = "respid"))
```

Now we are ready to prepare heat-maps.

## Heatmap for a single respondent-screen

Let's start with drawing a heat-map for a single respondent-screen. In the code chunk below screen at which set of question regarding immigrants were shown is selected and a single respondent is sampled.

```{r}
respCursorPositions <- cursorPositions %>%
  filter(screen == "immig") %>%
  filter(respid %in% sample(unique(respid), 1))
```

Having the data selected you may use *ggplot2* functions to plot a heat-map. Please note, that *relative* positions are used instead of *regular* (*absolute*) ones.

```{r}
# declaring data and mapping of variables onto axis
ggplot(respCursorPositions,
       aes(x = pageX_rel, y = pageY_rel)) +
  # heat-map itself
  geom_hdr() +
  # reversing the Y axis because of the difference in coordinates system
  # between web browsers and plotting packages
  scale_y_reverse() +
  # a decoration - marking the area within which INPUT elements were shown
  geom_vline(xintercept = c(0, 1), color = "white", linetype = "dashed") +
  geom_hline(yintercept = c(0, 1), color = "white", linetype = "dashed")
```

The darkest areas indicate no hovering and the lighter it gets, the more time was spent by the cursor over a given area. Please note that values in the legend indicate density, not simply the time (i.e. density is proportional to time, but given in a different unit).

## Several heatmaps on one graph

Let's suppose we want to compare heat-maps for several respondents on one graph. In the code chunk below 4 respondents with explicitly specified ids are selected.

```{r}
respsCursorPositions <- logData$actions %>%
  filter(screen == "immig") %>%
  filter(respid %in% c(988464343, 987844729, 988421695, 988555104))
```

To put heat-maps describing different respondents on separate panels of the graph, you need to add `facet_wrap()` call - see the last line of the code chunk below.

```{r}
# declaring data and mapping of variables onto axis
ggplot(respsCursorPositions,
       aes(x = pageX_rel, y = pageY_rel)) +
  # heat-map itself
  geom_hdr() +
  # reversing the Y axis because of the difference in coordinates system
  # between web browsers and plotting packages
  scale_y_reverse(limits = c(NA, -1)) +
  # a decoration - marking the area within which INPUT elements were shown
  geom_vline(xintercept = c(0, 1), color = "white", linetype = "dashed") +
  geom_hline(yintercept = c(0, 1), color = "white", linetype = "dashed") +
  # placing each respondent on a separate panel
  facet_wrap(vars(respid))
```

## Heatmaps comparing groups of respondents

Comparing group of respondents using heat-maps is also simple. You need to provide a data frame including all the respondents you want to compare between as data to the `ggplot()` call and then use respondent's characteristics as *faceting* variables. In the code chunk below function `facet_grid()` is used instead of `facet_wrap()` to arrange results in a grid constructed from crossing values of variables *gender* and *variant* (this second variable describes whether question about immigration was presented in the first or in the second block of the questionnaire).

As in the case of comparing specific respondents you have to choose whether you are interested in observing which group spent more time over a given region (then use `contour_var = "density"`) or which regions were more and which less *popular* within different groups (then use `contour_var = "ndensity"`). However, in case of larger groups of respondents differences are typically not so sharp here, as in case of specific respondents.

```{r}
# density
# declaring data and mapping of variables onto axis
ggplot(cursorPositions %>%
         filter(screen == "immig"),
       aes(x = pageX_rel, y = pageY_rel)) +
  # heat-map itself
  geom_hdr() +
  # reversing the Y axis because of the difference in coordinates system
  # between web browsers and plotting packages
  scale_y_reverse(limits = c(NA, -1)) +
  # a decoration - marking the area within which INPUT elements were shown
  geom_vline(xintercept = c(0, 1), color = "white", linetype = "dashed") +
  geom_hline(yintercept = c(0, 1), color = "white", linetype = "dashed") +
  # defining that graphs should summarize results
  # grouped by the crossing of variant and gender
  facet_grid(vars(variant), vars(gender))
```

# Drawing cursor traces and clicks

Sometimes you want to illustrate the cursor trace for a specific respondent-screen. Such a graph will look more informative if you enhance it by marking also position of clicks and if you place position of the survey form INPUT elements in the background. Consequently, you need to prepare few distinct data sets:

```{r}
# some single respondent's id
respondent <- 988543469
# data on cursor trace
respCursorMoves <- logData$actions %>%
  filter(screen == "immig",
         type %in% "mousemove") %>%
  filter(respid %in% respondent)
# data on clicks
respCursorClicks <- logData$actions %>%
  filter(type %in% "click") %>%
  semi_join(respCursorMoves,
            by = c("respid", "screen"))
# data on INPUT elements' positions
respInputPositions <- logData$inputPositions %>%
  semi_join(respCursorMoves,
            by = c("respid", "screen"))
```

Having the data prepared you may use *ggplot2* to prepare a graph:

```{r}
# no data specified, as it will be defined separately for each 'layer' of the plot
ggplot(mapping = aes(x = pageX_rel, y = pageY_rel)) +
  # putting INPUT elements into background
  geom_point(data = respInputPositions,
             size = 4, color = "darkgrey") +
  # drawing mouse trace
  geom_path(data = respCursorMoves,
            linewidth = 1.2) +
  # drawing clicks
  geom_point(data = respCursorClicks,
             size = 5) +
  # labeling start and end point
  geom_label(data = respCursorMoves[c(1, nrow(respCursorMoves)), ] %>%
               mutate(label = c("start", "end")),
             mapping = aes(label = label), alpha = 0.7) +
  # reversing the Y axis because of the difference in coordinates system
  # between web browsers and plotting packages
  scale_y_reverse()
```

# Animating cursor traces and clicks

If you want to animate cursor trace, it is possible by using package *gganimate* (you will also need a package responsible for *rendering* an animation - in this case this will be *gifski*, that prepares a GIF file, although this is not the only possible choice).

```{r}
library(gifski)
library(gganimate)
```

To prepare animation you need to assign a *ggplot* graph to an object:

```{r}
g1 <- ggplot(NULL,
       aes(x = pageX_rel, y = pageY_rel)) +
  # putting INPUT elements into background
  geom_point(data = respInputPositions,
             size = 4, color = "darkgrey") +
  # drawing mouse trace
  geom_path(data = respCursorMoves,
            linewidth = 1.2) +
  # drawing clicks
  geom_point(data = respCursorClicks,
             size = 5) +
  # labeling start and end point
  geom_label(data = respCursorMoves[c(1, nrow(respCursorMoves)), ] %>%
               mutate(label = c("start", "end")),
             mapping = aes(label = label), alpha = 0.7) +
  # reversing the Y axis because of the difference in coordinates system
  # between web browsers and plotting packages
  scale_y_reverse()
```

Now you may turn this graph into an animation by *adding* type of the animation in the same way as if you will add another graph element. In the case of drawing a cursor trace function `transition_reveal()` will provide the adequate type of animation and while calling it you need to specify that variable *timeStampRel* describes time:

```{r}
a1 <- g1 + transition_reveal(timeStampRel)
```

Now you need to *render* animation by calling function `animate()`:

```{r}
animate(a1, renderer = gifski_renderer())
```

Sadly, not everything works perfectly - while cursor trace itself is drawn correctly, the other *layers* (clicks, start and end labels) are shown in an odd way. For the moment I have not found the way to make *gganimate* works correctly while animating graph with many different *layers* that use different data. So the only solution to get rid of these strange effects is to animate a simpler version of the graph:

```{r}
a2 <- ggplot(NULL,
       aes(x = pageX_rel, y = pageY_rel)) +
  # putting INPUT elements into background
  geom_point(data = respInputPositions,
             size = 4, color = "darkgrey") +
  # drawing mouse trace
  geom_path(data = respCursorMoves,
            size = 1.2) +
  # reversing the Y axis because of the difference in coordinates system
  # between web browsers and plotting packages
  scale_y_reverse() +
  # make it an animation
  transition_reveal(timeStampRel)
animate(a2, renderer = gifski_renderer())
```
